Synthesising Complex Termination Arguments

### 2014-06-18 papers/termination/paper.tex

Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
over fixed-width binary words, otherwise known as bit-vectors.

%Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
%calulcations are just defined to 
The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
incomparable behaviours with respect to termination. 
%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
%% =======
%% most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
%% Next, we will consider in more detail the assumptions made by such techniques and their implications.\\

%% \noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
%% Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
%% integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
%% capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
%% Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
%% over fixed-width binary words, otherwise known as bit-vectors.

%% Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
%% calulcations are just defined to

%% The abstraction of bit-vectors to mathematical integers 
%% ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%% %{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
%% incomparable behaviours with respect to termination:
%% %The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
%% \begin{itemize}
%% \item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
%% \begin{lstlisting}[language=C]
%% while (x > 0) x -= 2;
%% \end{lstlisting}
%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
%% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
Computers tend to have fixed width integers.  For a machine with $k$-bit words,
all arithmetic is done modulo $2^k$.  This means that a program's termination
behaviour can depend strongly on whether it is interpreted as operating over
mathematical integers, or over fixed-width machine words.  We can however
observe that if no arithmetic operation \emph{overflows} (i.e. does not result
in a value greater than $2^k$), fixed-width and integer arithmetic coincide.

>>>>

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.}
The abstraction of bit-vectors to mathematical integers ignores
the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in
incomparable behaviours with respect to termination.

Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} underflows before becoming negative and goes back to being positive causing the loop to never terminate.

Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
Physical computers have bounded storage, which means they are unable to perform calculations on mathematical
integers.  For example, if $A$ is the Ackermann function and $G$ is Graham's number, a physical computer
capable of computing $A(G, G)$ would contain (much!) more matter than is believed to exist in the universe.
Fortunately, it is rare for a programmer to need such a large number and so modern computers do their arithmetic
over fixed-width binary words, otherwise known as bit-vectors.  For the remainder of this section, we will say that
the bit-vectors we are working with are $k$-bits wide, which means that each word can hold one of $2^k$ bit patterns.
Typical values for $k$ are 32 and 64.

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many of the differences between
machine arithmetic and Peano arithmetic.  To give an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.
This provides a counterexample to the statement $\forall x . x + 1 > x$, which is a theorem of Peano
arithmetic but not of modular arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its machine-arithmetic result is the same
as the result of the same operation performed on integers.

Machine words can be interpreted as ``signed'' or ``unsigned'' values.  Signed values can be negative,
while unsigned values cannot.  The encoding for signed values is two's complement, where the most significant
bit $b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather than $2^k - 1$.  Two's complement
representation has the property that $\forall x . -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
is bitwise-negation.  Two's complement also has the property that addition, multiplication and subtraction are defined
identically for unsigned and signed numbers.  Therefore signed and unsigned arithmetic differ only on comparison
operators\footnote{There is also the difference that signed overflow in a C program results in undefined behaviour,
but in practice the undefined behaviour is implemented just as if the arithmetic had been unsigned}.
For example: $$\forall x . -x \leq_s x \wedge -x \geq_u x$$ where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons respectively.

The final source of difference between integer arithmetic and bitvector arithmetic stems from width conversions.
Typical computers allow variables to have different types, which can be represented using words of different widths.
In C a \texttt{short} might occupy
16 bits, while an \texttt{int} might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit variable
with $l < k$, the result is truncated $\mod 2^l$.  For example, if $x$ is a 32-bit variable and $y$ is a 16-bit
variable, $y$ will hold the value $0$ after the following code is executed:

\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

This gives us a counterexample to the statement $\forall x, y . x = y \Rightarrow (x + 1) = (y + 1)$,
which is a theorem of Peano arithmetic.

As well as machine arithmetic differing from Peano arithmetic on the operators they have in common,
computers have several ``bitwise'' operations that are not taken as primitive in the theory of
integers.  These operations include the standard boolean operations \texttt{and, or, not, xor}
applied to each element of the bit-vector.  Computer programs often make use of these operators,
which are non-linear when interpreted in the standard model of Peano arithmetic
(although some of these operators can be seen as linear in a different algebraic structure,
e.g. \texttt{xor} corresponds to addition in the Galois field $\mathrm{GF}(2^k)$).
